Hijacking Simulators with Universal Probabilistic Programming

### 2019-05-21 paper/neurlps2019/main.tex
updated

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way. Thus, by design, simulators are ideal programs for probabilistic programming systems~(PPSs).
Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the PPS. 
% via the raw random number calls of the simulator.
 However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology , multi-agent and financial modeling. 
We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.

\todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}

@@

\section{Introduction}
\label{sec:intro}
\bg{The audience is a probabilistic programming audience - so that is what we will target}

@@

% \end{itemize}
\begin{wrapfigure}{R}{0.45\textwidth}
\centering
% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
\caption{Hijacking the OpenMalaria simulator}
\label{fig:hijacking} 
\end{wrapfigure}

Probabilistic programming\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
be viewed as the joint density of the user-specified model.
Having direct access to the source code enables probabilistic programming systems~(PPSs)
to conduct several processes in an automated fashion, such as 
the automation of complex inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}.
In addition to this, as models can be directly expressed in program code we can encode other structures to make use of the full structure of a given language,
which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.

One powerful \emph{class} of models are simulators. 
Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
Thus, Simulators by their very nature are
information rich and naturally describe complicated joint densities.
However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the PPS, which 
in many cases is not feasible, nor efficient due to both the size 
and complexity of the simulator code base. 
However, what if one could arbitrary connect a simulator, written in any
language and connect it directly to a PPS, without having to re-write
the simulator in the given PPS?

Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population-based simulators which arise
in a myriad of domains \bg{add citations}. 
Our framework provides a simple solution and makes it easy to transform arbitrary stochastic population-based simulators into probabilistic programs, regardless of the complexity of the simulator. 
Our system supports simulators written in 13 different programming languages and enables the user to perform for posterior inference over the density of the simulator. 
This enables two things.
1) It enables software developers and researchers to understand complicated code bases. 
2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome. 
This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

This allows for a variety of tasks to be performed on
the hijacked simulator, such as running inference (by conditioning
the values of certain draws and manipulating others),
uncovering stochastic structure, and automatically
producing result summaries, such as establishing the probability
of different program paths/traces.  By providing a common abstraction 
framework for different simulators, our approach further allows for
easy and direct comparison between related or competing
simulators, a characteristic that is valuable in the context of
epidemiology simulators \cite{ferris_openmalaria_2015}. We provide a case study of the above in Section \ref{sec:casestudy}.

Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible. This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

In time, we hope our approach will play a critical role in
bringing recent advancements in probabilistic programming to bear on
the vast array of existing simulators used throughout the sciences,
thereby providing wide-ranging impacts across a number of fields.

@@

\section{Background}
\label{sec:backgorund}

\begin{itemize}
\item Paragraph on Probabilistic programming
\item Paragraph on Simulators keep high-level focus on pop sims later
\item Paragraph on Baydin et al
\end{itemize}
In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
Recent advances in hardware have enabled simulations to model increasingly complex systems.
Epidemiology studies the prevalence and spreading of diseases across populations. Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.

Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
can be used to express probabilistic models and consequently perform automated inference these. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.

@@

\section{Hijacking Simulators}
\label{sec:methods}

@@

We first demonstrate our system on a reduced version of the

>>>>

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way. Thus, by design, simulators are ideal programs for probabilistic programming systems~(PPSs).
Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the existing PPS. 
% via the raw random number calls of the simulator.
 However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used within a probabilistic programming framework.
In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology, multi-agent and financial modeling and provide the relevant
theoretical underpinnings. 
We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.

% \todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}

@@

\section{Introduction}
\label{sec:intro}
% \bg{The audience is a probabilistic programming audience - so that is what we will target}

@@

% \end{itemize}

% One powerful \emph{class} of models are simulators. 
Simulators are useful tools when conducting experiments in the real system 
is impossible, or impractical and they arise in numerous industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008}. 
They are rich models and provide a level of interpretability not typically found in modern deep learning methodologies due to the several years, if not decades, of research and development that has led to their construction.
Thus, simulators by their very nature are information rich and naturally describe complicated joint densities.
That is the \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors $p(x)$. 
But, for all the benefits of using simulations there exist two challenging problems that are currently faced:
\begin{itemize}
\item It is difficult to calibrate the internal simulator parameters, as the events they are simulating are expensive to reproduce, or are very rare, so data is scarce. 
\item As the development process takes place over several years the code base becomes convoluted, making it increasingly challenging to understand the structure of the joint density and the internal operations within the simulator, diminishing the interpretability that the simulator once provided. 
\end{itemize}

\begin{wrapfigure}{R}{0.44\textwidth}
\centering
% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
\caption{An example of the posterior output of population-based simulator after hijacking. \textit{Top}: Represents the posterior execution traces of the simulator, inside the the probabilistic programming 
system. Each address $A_{i}$ correspond to a physical process within the simulator, we describe these addresses and their structures in more details in Section~\ref{sec:casestudy}.
\textit{Bottom}: Represents the histograms generated after posterior inference for the parameters of interest, in this case those parameters are represented by addresses $A_{2}$ to $A_{6}$, conditioned on some observations. 
The grey dashed lines represent the ground truth.}
\label{fig:hijacking} 
\end{wrapfigure}
One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,le2016inference}. 
Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}, which enables us to efficiently calibrate
model parameters, and we can also track how the internals of the program change during
run time, enabling a full analysis of the program structure and traces generated, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
any other language,
which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.

For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
This in many cases is not feasible, nor efficient due to both the size 
and complexity of the existing simulator code base. 
However, what if one could connect arbitrary a simulator, written in any
language and directly connect it to a PPS, without having to re-write
the simulator in the given PPS?

\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient}. 
They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions 
between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics  \bg{add citations}. 
Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator. 
Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including 
simulator output. We provide a theoretical justification for our approach and many examples.
% This enables two things.
% 1) It enables software developers and researchers to understand complicated code bases. 
% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome. 
% This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

% In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

% In time, we hope our approach will play a critical role in
% bringing recent advancements in probabilistic programming to bear on
% the vast array of existing simulators used throughout the sciences,
% thereby providing wide-ranging impacts across a number of fields.

@@

% \section{Background}
% \label{sec:backgorund}

% \begin{itemize}
% \item Paragraph on Probabilistic programming
% \item Paragraph on Simulators keep high-level focus on pop sims later
% \item Paragraph on Baydin et al
% \end{itemize}
% In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
% In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
% Recent advances in hardware have enabled simulations to model increasingly complex systems.
% Epidemiology studies the prevalence and spreading of diseases across populations. Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.

% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
% can be used to express probabilistic models and consequently perform automated inference these. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.

@@

\section{Hijacking Simulators}
\label{sec:methods}
In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
Recent advances in hardware have enabled simulations to model increasingly complex systems.
Epidemiology studies the prevalence and spreading of diseases across populations. Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.

@@

We first demonstrate our system on a reduced version of the OpenMalria simulator, proposed by Smith et .al~\cite{smith2006relationship}. The following model directly models the relationship between the entomological inoculation rate and the force of infection for the \emph{plasmodium flaciprum} Malaria.
